Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-28 paper-1/manuscript.tex
local/global adaptation -> suppression

Thus, we have finally traced the greater response to A in the dev sequence to its roots, and revealed two complementary roles of TA: The frequent presentation of non-target stimulus B in the dev sequence and concomitant increased activity in a subset of neurons responding early to B causes \emph{local adaptation} by TA in this subset, leading to smaller network responses. By contrast, stimuli in the msc sequence are presented more sporadically, allowing local adaptation to decay, thus leading to larger network responses. This difference in non-target response sizes -- small in dev, large in msc -- is then reflected in network-wide \emph{global adaptation}, such that the network is more excitable to stimulation with A (or other non-standard inputs) in the dev sequence.

@@

To understand how STD affects deviance detection, we first examined the effects of TA in the full model, following the analysis on the ablated model laid out in the previous section. Starting at the cause of the reduced response in non-target trials, we show in Figure \ref{fig7}a the full model equivalent of Figure \ref{fig6}c and d. The contrasts (dev - msc) in the non-target response magnitude and in the resulting $V_{TA}$ levels at the start of B trials both retained a clear separation between the portion of neurons responding early, which showed increased responses (T = 5.44e+03, p = 1.02e-06; one-sided Wilcoxon, n = 120) and increased TA levels (T = 3.41e+03, p = 9.51e-08), and the neurons responding later, whose response (T = 282, p = 9.19e-19) and TA levels (T = 72, p = 2.49e-17) decreased in the dev sequence. Likewise, the correlation across excitatory neurons between non-target response contrast and the contrast in $V_{TA}$ before B trials remained positive (T = 7.26e+03, p = 9.86e-22) and very strong (median Pearson's $\rho$ = 0.992) This indicates that local adaptation to B mediated by TA remained a key driver of the reduced response to non-target trials.

@@

Taken together, the data indicate that, with STD added to the model, the overall chain of causality -- increased local adaptation to B resulting in reduced B responses and reduced global adaptation, which in turn allows greater target responses -- remained intact with only minor quantitative changes. The primary cause for the greater target response, therefore, appeared to be the smaller non-target response, which was where we noted the first clear shift towards increased deviance detection.

@@

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than the non-target msc responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences.

>>>>

Thus, we have finally traced the greater response to A in the dev sequence to its roots, and revealed two complementary roles of TA: The frequent presentation of non-target stimulus B in the dev sequence and concomitant increased activity in a subset of neurons responding early to B causes \emph{local suppression} by TA in this subset, leading to smaller network responses. By contrast, stimuli in the msc sequence are presented more sporadically, allowing local suppression to decay, thus leading to larger network responses. This difference in non-target response sizes -- small in dev, large in msc -- is then reflected in network-wide \emph{global suppression}, such that the network is more excitable to stimulation with A (or other non-standard inputs) in the dev sequence.

@@

To understand how STD affects deviance detection, we first examined the effects of TA in the full model, following the analysis on the ablated model laid out in the previous section. Starting at the cause of the reduced response in non-target trials, we show in Figure \ref{fig7}a the full model equivalent of Figure \ref{fig6}c and d. The contrasts (dev - msc) in the non-target response magnitude and in the resulting $V_{TA}$ levels at the start of B trials both retained a clear separation between the portion of neurons responding early, which showed increased responses (T = 5.44e+03, p = 1.02e-06; one-sided Wilcoxon, n = 120) and increased TA levels (T = 3.41e+03, p = 9.51e-08), and the neurons responding later, whose response (T = 282, p = 9.19e-19) and TA levels (T = 72, p = 2.49e-17) decreased in the dev sequence. Likewise, the correlation across excitatory neurons between non-target response contrast and the contrast in $V_{TA}$ before B trials remained positive (T = 7.26e+03, p = 9.86e-22) and very strong (median Pearson's $\rho$ = 0.992) This indicates that local suppression to B mediated by TA remained a key driver of the reduced response to non-target trials.

@@

Taken together, the data indicate that, with STD added to the model, the overall chain of causality -- increased local suppression to B resulting in reduced B responses and reduced global suppression, which in turn allows greater target responses -- remained intact with only minor quantitative changes. The primary cause for the greater target response, therefore, appeared to be the smaller non-target response, which was where we noted the first clear shift towards increased deviance detection.

@@

This naturally raises the question of why, with the addition of STD, the non-target dev response was reduced more than the non-target msc responses. We hypothesized that, due to its relatively short time constant, the direct effect of STD across trials was limited to repeated trials of the same stimulus, such as the frequent stimulus in oddball sequences. This would mean that STD is capable only of local suppression, reducing the standard response, but not of global suppression, leaving networks with only STD unable to differentiate between dev and msc target trials.
